Improvement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-06-09 rigid_bodies.tex
files added

\subsection{Refinement of the NVT Dynamics}
\label{sec:refined_method}

@@

\section{Application to the Simulation of Liquid Water}
\label{sec:numerical_results}

In this section, we analyze the performance of our proposed NVT-MD method and compare its results to those obtained by employing both the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} and the method of \citeauthor{Martyna_1996} \cite{Martyna_1996} with our described adaptations.

The system under study has periodic boundary conditions and contains 903 TIP3P \cite{Jorgensen_1983} water molecules at a density of 970~kg/m\textsuperscript{3}.
The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at 10 {\AA} and smoothed by a switching function over the range from 9.5 \AA to 10 \AA, such as done in Ref.~\citenum{Silveira_2017}. 
The damping of Coulombic interactions was done by a factor $\text{erfc}(-\alpha r)$, with $\alpha = 0.29~\text{\AA}^{-1}$.
In addition, all thermostated simulations were carried out considering a time-scale constant $t_d = 100~\text{fs}$.

We start by examining the departure of the \textit{refined} Hamiltonian from the specified one in NVE simulations.
This is the most direct way of determining the importance of the discretization errors \cite{Engle_2005}, which is essential for discussing the NVT case.
In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system using different time-step sizes ($\timestep$).
For this, we used the numerical scheme of Eq.~\eqref{eq:splitting NVE} with the unsplit solution of Ref.~\citenum{Silveira_2017} applied for the rotations.
The four-point differentiation formulas presented in Sec.~\ref{sec:refined_method} were employed for computing the refined kinetic energy.
As expected, $\refined{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep = 1~\text{fs}$, but rapidly departs from $\Ham{}$ when $\timestep$ increases.
It is worth remarking that $\refined{\Ham{}}$ corresponds to a simple and, most importantly, computationally inexpensive approximation of the truly conserved shadow Hamiltonian.
As a result, we observe increasing fluctuations in $\refined{\Ham{}}$ for $\timestep = 3~\text{fs}$ and $\timestep = 4~\text{fs}$.
This is an indication that
1) the employed four-point formulas do not provide derivatives with the required accuracy or
2) the high-order terms of the Baker-Campbell-Hausdorff series (see Appendix \ref{sec:rigid body shadow hamiltonian}) would have a non-negligible impact on the shadow Hamiltonian.

@@

With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for each time-step size the free energy difference per water molecule ($\Delta F/N$, circles) between the sampled and target states.
$\Delta F$ corresponds to the denominator of Eq.~\ref{eq:reweighting}.
As expected, the free energy is only negligible for $h$ = 1 fs.
In Fig.~\ref{fig:free_energy} we also depict the \textit{refined} temperature ($\refined{T}$, squares), calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the target temperature.
Indeed, for all the time-step sizes considered the agreement is excellent.

@@

In our previous work\cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because the rotational degrees of freedom are the fastest modes in the rigid model of water. 
We see that the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
The results of Fig.~\ref{fig:energy_partition} (b), which were obtained for the numerical scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (c), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions. Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
From Fig.~\ref{fig:energy_partition} it is clear that our method corrects that serious disruption and a small deviation is only observed for 7 fs.

\begin{figure*}
	\includegraphics{Figures/props.eps}
	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
	\label{fig:properties}
\end{figure*}

>>>>

\subsection{Adjustment of the NVT Dynamics}
\label{sec:refined_method}

@@

\section{Application to Water and Discussion}
\label{sec:numerical_results}

In the following, we analyze the performance of the NVT scheme proposed in  this paper and compare the results with those obtained employing the integrator due to \citeauthor{Martyna_1996} \cite{Martyna_1996} as well as the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005}

The system under study corresponds to 903 TIP3P\cite{Jorgensen_1983} water molecules at a density of 970 kg/m$^3$.
The 12-6 Lennard-Jones and damped Coulombic interactions were truncated at  10 {\AA} and smoothed via a switching function over the range of 9.5 {\AA} to  10 {\AA} as done in Ref.~\citenum{Silveira_2017}. 
The damping of Coulombic interactions was done by a factor erfc($\alpha r$), with $\alpha$ = 0.29 {\AA}$^{-1}$.
In addition, all the simulations were carried out considering $t_d = 100$ fs.

Before presenting our main results, we first examine the departure of the \textit{refined} Hamiltonian from the specified one, as this is the most direct way to determine the importance of the discretization erros \cite{Engle_2005}, which is essential for discussing the NVT case.
In Fig.~\ref{fig:nve} we show the time evolution of both $\refined{\Ham{}}$ (dotted lines) and $\Ham{}$ (solid lines) for the water system undergoing NVE dynamics with different time-step sizes ($\timestep$).
For this, we used the numerical scheme of Eq.~\ref{eq:splitting NVE} with the unsplit solution of rotation.
As expected, $\refined{\Ham{}}$ is very well conserved and close to $\Ham{}$ for $\timestep$ = 1 fs, but fairly rapidly separates from $\Ham{}$ when increasing $\timestep$. 
It should be remarked that $\refined{\Ham{}}$ corresponds to a simple approximation and, most important, it is computationally very cheap.
As a result of such approximation, we have larger fluctuations in $\refined{\Ham{}}$ for $\timestep$ = 3 fs and $\timestep$ = 4 fs (add in conclusions that we should use high-order approximations).

@@

With the aim of highlighting the importance of reweighting, in Fig.~\ref{fig:free_energy} we show for each time-step size the free energy difference per water molecule ($\Delta F/N$, circles) between the sampled and target states.
$\Delta F$ corresponds to the logarithm of the averaged denominator of Eq.~\ref{eq:reweighting}.
As expected, the free energy is only negligible for $h$ = 1 fs.
In Fig.~\ref{fig:free_energy} we also depict the \textit{refined} temperature ($\refined{T}$, squares), calculated by means of the generalized equipartition theorem, given by Eq.~\ref{eq:equipartition}.
Of course, as $\refined{T}$ is indirectly controlled by means of Eq.~\ref{eq:thermostat_force}, it is expected to reproduce the target temperature.
Indeed, for all the time-step sizes considered the agreement is excellent.

@@

In our previous work\cite{Silveira_2017}, we demonstrated that in the NVE case the rotational kinetic energy is more influenced by the time-step size because the rotational degrees of freedom are the fastest modes in the rigid model of water. 
We see that the method of \citeauthor{Kamberaj_2005} \cite{Kamberaj_2005} reproduces the behavior of the NVE case, as noted before.
The results of Fig.~\ref{fig:energy_partition} (b), which were obtained for the numerical scheme of \citeauthor{Martyna_1996} \cite{Martyna_1996}, are similar to Fig.~\ref{fig:energy_partition} (c), in the sense that the simulated translational and rotational kinetic energies deviate almost symmetrically and in opposite directions. Also, they do so contrary to what is expected: the translational kinetic energy should be smaller than the rotational one.
From Fig.~\ref{fig:energy_partition} it is clear that our method corrects that serious disruption for time-step sizes up to 5 fs.

\begin{figure*}
	\includegraphics{Figures/thermodynamic_properties.eps}
	\caption{Effect of the time-step size on the thermodynamic properties and of 903 TIP3P\cite{Jorgensen_1983} water molecules in NVT MD simulations employing...}
	\label{fig:properties}
\end{figure*}
